% pdflatex
\documentclass[12pt]{article}
\usepackage{ebgaramond}
\usepackage{microtype}

\newcommand{\vin}{\hspace{1em}}

\title{The Story of \textsc{Nefrekepta}}
\author{Gilbert Murray\footnote{G.~Murray, \textit{The Story of Nefrekepta,
From a Demotic Papyrus}, 1911.}}
\date{}

\begin{document}
\maketitle


\section*{Preface}

The original of this tale is in Demotic Egyptian, in prose, on a fragmentary
papyrus dated ``the first month of winter, in the fifteenth year'' of some king
unnamed. Palaeographical evidence suggests some date about 100 \textsc{bc}. My
own education has been neglected in the matter of Demotic, and I know the tale
only from the literal translation which accompanies the text in Dr.~Griffith's
\textit{Stories of the High Priests of Memphis}. In that form, however, it so
fascinated me that I presently found myself, to the neglect of more urgent
duties, putting it into English verse and filling up the gaps in the narrative.
I have tried to preserve the style and often the exact words of the original,
as rendered by Dr.~Griffith, but in other respects temptations have been great
and I have not resisted them.

The names present some difficulties. In Demotic, if I understand aright, the
vowels are not written and the consonants often do not mean what they seem to
mean. The hero's name is spelt, roughly speaking, N('y)-nfr-k'-Pth, the phantom
lady's Ty-bwbwe; the priest's is written Stne, but was probably pronounced, so
Dr.~Griffith tells me, S\v{e}t\^{o}n or S\v{e}t\^{a}n.

While doing the verses I was constantly reminded of certain Egyptian
illuminations by Miss Florence Kingsford, now Mrs.~Sydney Cockerell, which I
had seen some years ago, and she has been so kind as to provide the book with a
frontispiece and tail-piece.\footnote{These illustrations are not reproduced in
this edition.}

Those who desire further information about \textsc{Setne} will find it in the
introduction to Dr.~Griffith's learned and delightful book (Clarendon Press,
1900).

\null\hfill G.~M.


\section*{Introduction}

\begin{verse}
\textsc{Setne Khamuas}, son of \textsc{Rameses},\\*
High Priest of \textsc{Ptah}, beneath his garden trees\\*
\vin Dwelt with his wife and children; wise was he\\*
In Books of power and ancient Masteries.

And much he pondered on a tale they told,\\*
How \textsc{Nefrekepta}, dead in days of old,\\*
\vin Held still the Book of \textsc{Hermes} which is \textsc{Thoth}\\*
Hid in his tomb, and never loosed his hold.

And longing for that Book so pierced him through\\*
He called to him his brother \textsc{An-Herru},\\*
\vin The son of \textsc{Menkh-Art}, saying: ``Brother mine,\\*
Be with me in a deed I have to do.''

And he said: ``I am with thee till my doom\\*
Find me.'' And \textsc{Setne} said: ``I seek the tomb\\*
\vin Of \textsc{Nefrekepta} dead, and take the Book,\\*
The Book of \textsc{Thoth} hid in its inmost room.''

That night they found the tomb, and \textsc{An-Herru}\\*
Stayed at the door, but \textsc{Setne} passing through\\*
\vin On seven great doors and seven windings spake\\*
His spells, and found the room; and all was true.

For there lay \textsc{Nefrekepta} in his pride,\\*
The Book beneath his head; and at his side\\*
\vin The ghosts sate of a woman and a boy,\\*
Shadows beside the dead; and \textsc{Setne} cried:

``Ye thron\`{e}d Shadows, whosoe'er ye be,\\*
And thou, dead \textsc{Pharaoh}, tombed in majesty,\\*
\vin All hail! I, \textsc{Setne}, scribe and Priest of \textsc{Ptah},\\*
Command thy Book be rendered up to me,

The Book of \textsc{Thoth} which lies beneath thy head.''\\*
Then never word nor sign came from the dead,\\*
\vin But the two Shadows lifted up their arms\\*
Lamenting, and the woman swift outspread

Her hand to save the Book, and cried: ``Aha!\\*
\textsc{Setne Khamuas}, is the will of \textsc{Ra}\\*
\vin Not yet fulfilled upon us? This is I,\\*
\textsc{Ahure} of the race of \textsc{Merneb-Ptah}.''
\end{verse}


\section*{\textsc{Ahure}'s Story}

\begin{verse}
``Therefore give ear, and let the Book of Gold\\*
Tempt thee no more, till all my tale be told.---\\*
\vin It happened in the days of \textsc{Merneb-Ptah},\\*
The long days, when the King was very old,

And had no son nor daughter; in their room\\*
A son's son and a daughter's daughter, whom\\*
\vin Thou seest, \textsc{Nefrekepta} the Good Scribe,\\*
And me who watch beside him in this tomb.

And \textsc{Pharaoh} mused and spake: `Go near and far,\\*
Bring me the lordliest of my chiefs of war\\*
\vin With all their daughters and their sons, to make\\*
Feast on the third day; call me all that are.'

So \textsc{Pharaoh} spake; but lo, exceedingly\\*
Did I love \textsc{Nefrekepta} and he me.\\*
\vin And much I feared that \textsc{Pharaoh} at that feast\\*
Would take some youth and maid of high degree,

A war-lord's daughter and a war-lord's son,\\*
And fast bind \textsc{Nefrekepta} to the one,\\*
\vin And me to the other; so his race would grow,\\*
But we two meet no more beneath the sun.

Now \textsc{Pharaoh}'s ancient Steward loved me well,\\*
And \textsc{Nefrekepta} too; and it befell,\\*
\vin One day he watched me and the second day\\*
He spoke: `\textsc{Ahure}, surely I can spell

A story that is written in four eyes.\\*
Thou lovest \textsc{Nefrekepta} and likewise\\*
\vin He loveth thee.' And I cried out: `O friend,\\*
Speak unto \textsc{Pharaoh} quick, ere this day dies!

Pray that he give me to my cousin straight,\\*
Nor seek to make us two live separate.'\\*
\vin And he said: `I will speak; for so the law\\*
Commandeth, by long ages consecrate;

The King's sons wed the daughters of the King.'\\*
Then all my heart was like a water-spring\\*
\vin Leaping; and soon he went and soon returned\\*
Sad, and reported of his counselling:

`I spake to \textsc{Pharaoh}, saying: ``Lord, may life\\*
Like \textsc{Ra}'s be thine and glory in all strife!\\*
\vin Is it not meet that \textsc{Nefrekepta} take\\*
By ancient rule \textsc{Ahure} to his wife?

Let \textsc{Pharaoh} wed with \textsc{Pharaoh}; so shall pure\\*
\textsc{Pharaoh} be born:---the rule doth still endure.''\\*
\vin And \textsc{Pharaoh} spoke not, but his brow grew dark\\*
With trouble; and I said: ``O King, for sure

Thou hast some grief; say what doth vex thy brow?''\\*
And \textsc{Pharaoh} said: ``None vexeth me but thou.\\*
\vin I have but two, and if those two be wed\\*
All \textsc{Pharaoh}'s fruit is hanging on one bough.

Nay, search and find me one great war-lord's son,\\*
Another war-lord's daughter. With the one\\*
\vin And other let these children twain be wed.\\*
So \textsc{Pharaoh}'s line on many threads shall run.''{'}

Next day the hours passed and the feast was set\\*
Before the King; and I was called ere yet\\*
\vin The Lords came. And I stood before the King\\*
Not as of old, my heart being desolate.

And \textsc{Pharaoh} spake: ``\textsc{Ahure}, was it thou\\*
Didst send that message that hath made my brow\\*
\vin Troubled, that with thy brother thou wouldst wed?''\\*
And I said: ``Hath the King not made a vow,

A war-lord's daughter and a war-lord's son---\\*
Behold us!---Let my brother take the one,\\*
\vin And let the other take \textsc{Ahure}; so\\*
Shall \textsc{Pharaoh}'s race be great beneath the sun.''

And there I laughed; and \textsc{Pharaoh} laughed again,\\*
And called the Steward of the King's domain:\\*
\vin `Steward, this night to \textsc{Nefrekepta}'s house\\*
The maid \textsc{Ahure} take and all her train;

And all things beautiful go with her there.'\\*
Thus I to \textsc{Nefrekepta}'s house did fare,\\*
\vin And \textsc{Pharaoh} sent wrought silver and fine gold,\\*
And \textsc{Pharaoh}'s servants stood about my chair.

And \textsc{Nefrekepta} took with me delight,\\*
And feasted \textsc{Pharaoh}'s servants; and that night\\*
\vin Was made our marriage, and we knew great joy,\\*
And never, never, failed I in his sight;

For each his fellow loved exceedingly.\\*
And when my time of bearing came to be\\*
\vin I bore the son who lieth in this tomb,\\*
\textsc{Merab}; a name in the Kings' Book is he.''
\end{verse}


\section*{The Book of \textsc{Thoth}}

\begin{verse}
``And thus it fell, that of all things on earth\\*
My brother \textsc{Nefrekepta} most of worth\\*
\vin Did hold the wisdom that in Books is writ.\\*
The tablets of the House of Death and Birth,

And all that on the temple walls is said,\\*
And all the lore of the Kings' Tombs he read,\\*
\vin And ever walked in \textsc{Memphis} on the Hill\\*
Of Kings, and stored the wisdom of the dead.

Now one day was a high procession sent\\*
To \textsc{Ptah}'s great house; and \textsc{Nefrekepta} went\\*
\vin And walked therein; right slow he walked, and read\\*
All that was writ on wall and pediment.

And, watching him, behold, an aged Priest\\*
Laughed. And he said: `God's mercy be increased!\\*
\vin Why laughest thou at me?' And he: `I laugh\\*
At no man, not the greatest nor the least;

I laugh to think how thou shalt laugh anon,\\*
When that which no man's eye hath looked upon,\\*
\vin The secret Book of \textsc{Hermes} which is \textsc{Thoth},\\*
Is opened to thine hand and called thine own.

He wrote it, and the Gods before his face\\*
Fled.---And it lies... If ever word of grace\\*
\vin Or spell of power thou need, come then to me\\*
And speak. Thou shalt be guided to the place.

Two leaves it hath, on which two sorceries\\*
Are written. If thou read the first of these,\\*
\vin Thou shalt enchant the earth, the clouds above,\\*
The underworld, the mountains and the seas;

And all the words that wing\`{e}d things may say,\\*
And creeping things, shall be made thine that day;\\*
\vin Yea, thou shalt see all fishes in the deep\\*
And God's power guiding each upon his way.

And if thou read the second, though there lies\\*
Above thee all \textsc{Amenti}, thou shalt rise\\*
\vin And take thy shape again, and see the Moon\\*
And \textsc{Ra} and all the children of the skies.'

And \textsc{Nefrekepta} cried: `O Priest and King!\\*
I bless thee. Tell me every gorgeous thing\\*
\vin Thy soul desireth, they shall all be thine,\\*
Wilt thou but guide my steps to that great spring.'

Then smiled the Priest: `My Prince, so let it be!\\*
Send me an hundred bars of silver, free\\*
\vin Of all fault, for my burial when I die;\\*
And two full priesthoods give me without fee.'

And \textsc{Nefrekepta} called a youth, and bade\\*
An hundred bars of silver pure be made,\\*
\vin And two new priesthoods named to \textsc{Thoth} and \textsc{Ptah},\\*
And sealed him priest to \textsc{Ptah} and \textsc{Thoth} unpaid.

Then the man spoke: `The Book of \textsc{Thoth} doth rest\\*
In \textsc{Coptos} Sea, hid in a golden chest;\\*
\vin The gold doth lie in silver; that in wrought\\*
Ebon and ivory, fitted nest in nest;

That in sweet cedar; that in bronze doth lie;\\*
The bronze in iron. 'Tis knotted with a ply\\*
\vin Of endless Snake; and round it for one league\\*
Are scorpion, asp, and worm to make men die.'

He spoke, and \textsc{Nefrekepta} no more knew\\*
What place he stood in nor what breath he drew,\\*
\vin But forth he hied him in great joy, and caught\\*
My hand, and all this tale he told me true;

And cried: `South, South to \textsc{Coptos}! None shall stay\\*
Our going.' But I turned from him away\\*
\vin And found that aged Priest and said: `Thou Priest,\\*
May \textsc{Amun} curse thee for thy words this day!

The Book of \textsc{Thoth}, the serpents and the sea!\\*
Most bitter striving thou hast made for me,\\*
\vin And bitter watching till my lord's return;\\*
False art thou, and thy South all cruelty.'

And much I prayed my brother not to sail\\*
To \textsc{Coptos}, but my prayer might not avail,\\*
\vin For straight to \textsc{Pharaoh}'s throne he went, and spake\\*
To \textsc{Pharaoh} of the Priest and all his tale.

And \textsc{Pharaoh} said: `What wilt thou I should do?'\\*
`Give me thy ship of pleasure and its crew,'\\*
\vin He said: `and with me let \textsc{Ahure} sail\\*
To find the Book, and the boy \textsc{Merab} too.'

So \textsc{Pharaoh}'s pleasure-ship was brought, and all\\*
Its crew; and southward in high festival\\*
\vin We sailed to \textsc{Coptos}; and the news went forth\\*
Before us, and the folk stood on the wall.

The Priests of \textsc{Isis} and \textsc{Harpocrates}\\*
And the Chief Priest of \textsc{Isis}, all of these\\*
\vin Came down to \textsc{Nefrekepta}, and to me\\*
The women of the Priests in their degrees.

They led us to their Temple in a line;\\*
And \textsc{Nefrekepta} gave ox, goose, and wine,\\*
\vin And brought to \textsc{Isis} and \textsc{Harpocrates}\\*
Rite and oblation and all dues divine.

A temple beautiful exceedingly\\*
Was over us, and there four days did he\\*
\vin Make with the Priests of \textsc{Isis} holiday,\\*
And the priests' women holiday with me.

But when the morning of the fifth day came,\\*
He called for wax made holy, without blame,\\*
\vin And shaped a boat with men, and spake a spell,\\*
And breathed; and life woke in them like a flame.

He went on board that boat, and heaped it high\\*
With sand, and pushed it from the shore. And I\\*
\vin Sate on the shore alone, and said: `I wait\\*
Here till he comes, and if he dies I die.'

He said: `O Rowers, row me to the place\\*
I wot of.' And the rowers rowed apace,\\*
\vin By day, by night, and ceased on the third day.\\*
Then he took sand, and cast the sand a space

Before him, and the water rose on both\\*
Sides, and the floor below did ooze and froth\\*
\vin With scorpion, asp, and worm to make men die,\\*
One league before the treasure chest of \textsc{Thoth}.

And round the chest was coiled an endless Snake.\\*
Then \textsc{Nefrekepta} took his charms, and spake\\*
\vin A spell on all that league of serpent things,\\*
And down they sank, and slept, and could not wake.

Then o'er the league of asps he walked, and fought\\*
With the endless Snake and slew it; but it caught\\*
\vin Life as it fell, and joined again and flew\\*
To tear him; and again its death he wrought;

Then a third time it quickened, and again\\*
He fought and smote the endless Snake in twain\\*
\vin The third time; and between the parts he cast\\*
Fine sand; and it lay still, for ever slain.

And on he strode and found the caskets, rolled\\*
Each within each; iron the outmost fold,\\*
\vin Then bronze; then cedar; then came ivory\\*
And ebon; then the silver; then the gold.

He broke the gold, and kneeling on his knees\\*
Read out the first of the two sorceries,\\*
\vin Enchanting all the earth, the clouds above,\\*
The underworld, the mountains and the seas.

And everything that bird or worm might say,\\*
Or mountain beast, he heard and knew that day,\\*
\vin And saw all fishes moving in the deep,\\*
And God's power guiding each upon its way.

He read the second sorcery, and far\\*
In heaven he saw the shining forth of \textsc{Ra},\\*
\vin And all his Children round him, and the Moon\\*
Uprising, and the shape of every Star.

And \textsc{Nefrekepta} shut the Book, and then\\*
Went to his boat and called the magic men:\\*
\vin `Row day, row night, and row me to the shore.'\\*
So rowed they, and he reached the shore again.

And found me sitting by the sea to wait\\*
His coming. Seven full days and nights I sate,\\*
\vin And ate nor drank, but waited, and was grown\\*
Like them they bear to the embalmer's gate.

I said to him: `My brother, let me see\\*
The Book that wrought such pain on thee and me.'\\*
\vin He gave the Book into my hands, and slow\\*
I read what there was writ of sorcery.

I laid upon the earth and sky my spell,\\*
The underworld, the hills, the ocean swell;\\*
\vin And understood what tales the birds of heaven\\*
And mountain beasts and deep-sea fishes tell.

Then did I read the second spell, and high\\*
Above saw \textsc{Ra} enthron\`{e}d in the sky,\\*
\vin And all his Children; and I saw the Moon\\*
And all the Stars in all their shapes go by.

And \textsc{Nefrekepta} bade them bring a roll\\*
Of new papyrus, and wrote out the whole\\*
\vin Of those two charms, and melted it in wine,\\*
And drank it. So the charm was in his soul.

Then sacrifice we made and gifts of worth\\*
Heaped high in all the temples, and set forth\\*
\vin On \textsc{Pharaoh}'s ship with singing and great joy,\\*
One league, one league, from \textsc{Coptos} to the North.

But \textsc{Thoth} himself had seen our deed, and, ah,\\*
His wrath was hot! Before the throne of \textsc{Ra},\\*
\vin `Judgement!' he cried, `Give judgement between me\\*
And \textsc{Nefrekepta}, son of \textsc{Merneb-Ptah},

Who broke into my treasure-house this day,\\*
And slew my Snake and stole my Book away.'\\*
\vin And \textsc{Ra} said: `Surely he is in thy hand,\\*
O \textsc{Thoth}, both he and his, to spare or slay.'

And lo, a Power of God went forth, and fell\\*
On all the river and lay invisible;\\*
\vin And \textsc{Thoth} said: `\textsc{Nefrekepta} shall come home\\*
No more, nor one of those that with him dwell.'

Then the boy \textsc{Merab}, singing, from the shade\\*
Of \textsc{Pharaoh}'s awning stepped; one step he made,\\*
\vin And, lo, the River took him, and his face\\*
Was covered and the will of \textsc{Ra} obeyed.

Then all about us cried with a great cry.\\*
But \textsc{Nefrekepta} from his awning high\\*
\vin Called with a spell, and the dead boy rose up;\\*
But over him that Power of God did lie.

Then \textsc{Nefrekepta} spake a written spell,\\*
And the boy \textsc{Merab} told all that befell\\*
\vin About him; yea, the very words which \textsc{Thoth}\\*
Spake at the throne of \textsc{Ra} he made him tell.

Then slow to \textsc{Coptos} we returned, and bore\\*
The boy \textsc{Merab} to the embalmer's door;\\*
\vin And like a Prince he was embalmed and laid\\*
On \textsc{Coptos} Hill with the great dead of yore.

And \textsc{Nefrekepta} said: `My sister, come\\*
Quick; let us row till all the road be foam,\\*
\vin Lest \textsc{Pharaoh} hear what hath befallen us,\\*
And his heart faint because we come not home.'

We went on board, and northward rowed apace\\*
One league from \textsc{Coptos}; and I saw the place\\*
\vin Where \textsc{Merab} died, and from the canopy\\*
Stepped, and the River took me; and my face

Was covered and the will of \textsc{Ra} obeyed.\\*
Then all upon the boat great mourning made,\\*
\vin But \textsc{Nefrekepta} from his awning high\\*
Called, and I knew a spell upon me laid;

And I rose up, though o'er me still had hold\\*
That Power of God. I rose and did unfold\\*
\vin All that befell me, yea, and every word\\*
Which \textsc{Thoth} before \textsc{Ra}'s throne had spoke I told.

Then slow to \textsc{Coptos} he returned and bore\\*
Me, his dead sister, to the embalmer's door;\\*
\vin And like a Queen I was embalmed, and laid\\*
Where \textsc{Merab} my dead child was laid before.

He went on board, and down the stream apace\\*
Rowed one league north from \textsc{Coptos}, to the place\\*
\vin Where \textsc{Merab} died and I, \textsc{Ahure}, died;\\*
And stood and communed with his heart a space:

`Shall it be backward now, ere \textsc{Thoth} can slay?\\*
Shall it be on to \textsc{Pharaoh}, come what may?\\*
\vin And, lo, when \textsc{Pharaoh} asks me where those two,\\*
His children, are, what is it I shall say?

``I took thy children to a burning land\\*
And living let them die; and here I stand.''\\*
\vin I will not speak it.'---Then he bade them bring\\*
A band of finest linen, such a band

As dead kings wear, to bind them at the last;\\*
And seven times round his body made it fast,\\*
\vin And close against his body bound the Book\\*
Firm; and from out the canopy he passed.

And, lo, the River took him, and the will\\*
Of \textsc{Ra} was done. And they on board did fill\\*
\vin The air with wailing: `Great woe! Grievous woe!\\*
Dead, dead, is the Good Scribe and all his skill.'

And down the stream the pleasure-ship sailed on\\*
Toward \textsc{Memphis}, and to no man there was known\\*
\vin Where \textsc{Nefrekepta} lay; and when they came\\*
Message was brought to \textsc{Pharaoh} on his throne.

And \textsc{Pharaoh} came in robes of funeral,\\*
And all the folk of \textsc{Memphis}, great and small,\\*
\vin And \textsc{Ptah}'s High Priest and all the Priests of \textsc{Ptah},\\*
And \textsc{Pharaoh}'s council and his household, all;

And saw the ship, and, lo, beneath it drowned\\*
Saw \textsc{Nefrekepta} lying, both hands wound\\*
\vin About the rudder, guiding still his ship;\\*
So great a scribe was \textsc{Nefrekepta} found.

They raised him, and against his body dead\\*
They found the Book pressed close. And \textsc{Pharaoh} said:\\*
\vin `Behold the Book he died for! Let it lie\\*
In this King's grave, a pillow for his head.'

Then sixteen days embalming did they keep\\*
For \textsc{Nefrekepta}, thirty-five of deep\\*
\vin Wrapping; of burial threescore days and ten;\\*
And here he resteth in the House of Sleep.

And I, \textsc{Ahure}, far away must lie\\*
In \textsc{Coptos}; but my heart within doth cry\\*
\vin For \textsc{Nefrekepta}, and our shadows come\\*
Waking and watch beside him sleeplessly.''
\end{verse}


\section*{The Contest for the Book}

\begin{verse}
``And thou, \textsc{Setne Khamuas}, who dost look\\*
To take from us that which from \textsc{Thoth} we took,\\*
\vin Where hast thou paid the price?---These twain and I,\\*
Our lives on earth were taken for this Book.''

But \textsc{Setne} said: ``\textsc{Ahure}, none the less,\\*
For all thy tale of old unhappiness,\\*
\vin Yield me the Book at \textsc{Nefrekepta}'s head:\\*
I take it else by wrath and bitter stress.''

Then \textsc{Nefrekepta} from his marble bed\\*
Rose up: ``O thou to whom my wife hath said\\*
\vin Words vainly wise, and thou hast hearkened not,\\*
How wilt thou win the Book beneath my head?

Think'st thou to take it from me by the wit\\*
Of a good scribe? Or wilt thou play for it\\*
\vin Four games of draughts, the fifty points and two?''\\*
And \textsc{Setne} said: ``The draughts;'' and down did sit.

Between them then the gaming board they drew\\*
For the first game of fifty points and two;\\*
\vin And \textsc{Nefrekepta} won the game, and spake\\*
A spell, and with the board he ruled it true.

And \textsc{Setne} ankle-deep into the floor\\*
Sank. And again they played, and as before\\*
\vin Dead \textsc{Nefrekepta} won the second game;\\*
And \textsc{Setne} sank up to the thighs and more.

Again they played; and \textsc{Nefrekepta} dead\\*
Won the third game; and \textsc{Setne} to his head\\*
\vin Sank, and the earth came close below his ears.\\*
And \textsc{Setne} lifted up his voice in dread:

``Brother, where art thou? Brother \textsc{An-Herru},\\*
Bear word to \textsc{Pharaoh} of the deeds I do.\\*
\vin Pray \textsc{Pharaoh} the last amulets of \textsc{Ptah}\\*
Be sent me. Haste! Haste!''---Then the dead man drew

The board up close, and the fourth game began.\\*
And \textsc{An-Herru} far off had heard, and ran,\\*
\vin Ran unto \textsc{Pharaoh}'s throne and told his tale.\\*
And \textsc{Pharaoh} said: ``To save a sinking man

Take the last amulets!'' And \textsc{An-Herru}\\*
Flew with them to the dead man's gate, and flew\\*
\vin Through all the windings, all the doors, and, lo,\\*
The game was playing still between those two.

He laid the amulets on \textsc{Setne}'s head,\\*
And \textsc{Setne} shivered in the earth, and said\\*
\vin A great spell; then, upstriving from the ground,\\*
Reached out his arm, and caught the Book, and fled.

Light went before him, and behind great gloom\\*
Closed, and he heard \textsc{Ahure} for her doom\\*
\vin Wailing: ``King Darkness, come! King Light, farewell!\\*
Gone, gone, is the last comfort of the Tomb.''

But \textsc{Nefrekepta} on his marble bed\\*
Lay back and laughed: ``A little while,'' he said,\\*
\vin ``O Shadow of my Sister, and this man\\*
Shall come again. Therefore be comforted.

His dreams shall bring him back, before mine ire\\*
Kneeling, to do the worst of our desire,\\*
\vin A fork upon his neck, a rod between\\*
His hands, and on his head a bowl of fire.''
\end{verse}


\section*{\textsc{Ta-Buvu\"{e}}, and the Return of the Book}

\begin{verse}
But \textsc{Setne} out into the light above\\*
Returned, and saw the light with a great love;\\*
\vin And sealed the tomb, and stood at \textsc{Pharaoh}'s throne\\*
And told his doings and the end thereof.

And \textsc{Pharaoh} said to \textsc{Setne}: ``These be vain\\*
Doings. Go back and give the dead again\\*
\vin His book, as a wise man to a wise man,\\*
Now, lest thou give it some day with much pain.''

But \textsc{Setne} hearkened not. By day and night\\*
He read the Book and took therein delight,\\*
\vin And showed it at his feasts; and all his days\\*
Were sweet to \textsc{Setne} and his breath was light.

'Twas joy to read, joy also when he made\\*
Mirth with his sons, joy when he rose and prayed\\*
\vin In \textsc{Ptah}'s great Temple; till one day, behold,\\*
In \textsc{Ptah}'s great Temple, through the colonnade,

A troop of damsels fair exceedingly,\\*
And one who led them. Beautiful was she,\\*
\vin And not like other women; good beyond\\*
All he had seen or ever thought to see.

Her girdle was of gold and gold her hood,\\*
And all that touched her fragrant was and good,\\*
\vin And maids behind her fifty walked and two;\\*
And, seeing, \textsc{Setne} wist not where he stood,

But called the Slave that served him: ``Haste thee, go\\*
To where yon woman worships; I would know\\*
\vin What need hath brought her and what name she bears.\\*
Go swiftly.'' And the servant bowed him low,

And found a handmaiden who walked aside:\\*
``Damsel, thy mistress cometh in much pride;\\*
\vin Say by what name men call her, and what need\\*
Brings her to \textsc{Memphis}.'' And the maid replied:

``This is the child of the Chief Vision Seer\\*
Of \textsc{Bast}, Queen of the Far World and the Near;\\*
\vin Her name is \textsc{Ta-Buvu\"{e}}, and a vow\\*
To \textsc{Ptah}, your mighty God, hath brought her here.''

The slave returned and told to \textsc{Setne} all\\*
The handmaid spoke; and \textsc{Setne} said: ``Go, call\\*
\vin This handmaid secretly behind the rest;\\*
Greet her from me and speak my name withal:

`\textsc{Setne Khamuas}, son of \textsc{Rameses},\\*
Greets thee: he sends ten gold Arsino\"{e}s;\\*
\vin And more, if any man hath done thee wrong,\\*
\textsc{Setne} the judge will right thine injuries;

All this, if thou wilt speak with him an hour,\\*
And help him. To a great and secret tower\\*
\vin Thou shalt go in, where none shall do thee hurt\\*
Nor know thy name; so great is \textsc{Setne}'s power.'{''}

The slave returned and all of \textsc{Setne}'s word\\*
Told to the handmaid; and her wrath was stirred,\\*
\vin And loud she railed, as though 'twere blasphemies\\*
His lips had spoke; and \textsc{Ta-Buvu\"{e}} heard,

And called him: ``Strive not with this foolish one;\\*
But hither, tell to me what wrong is done.''\\*
\vin But quick the handmaid ran before, and cried:\\*
``He bringeth words of shame from \textsc{Pharaoh}'s son:

Thus saying: `\textsc{Setne}, son of \textsc{Rameses},\\*
Greets thee, and sends ten gold Arsino\"{e}s;\\*
\vin And more, if any man hath done thee wrong,\\*
\textsc{Setne} the judge will right thine injuries;'

All this, if I will speak with him an hour\\*
And help him. To a great and secret tower\\*
\vin I shall go in, where none shall do me hurt\\*
Nor know my name; so great is \textsc{Setne}'s power!''

Then \textsc{Ta-Buvu\"{e}} laughed: ``I think he spake\\*
This word to thee for \textsc{Ta-Buvu\"{e}}'s sake.---\\*
\vin Go, speak to \textsc{Setne}, saying: `Who am I\\*
That thou shouldst send my bondmaid gifts to take?

I am no common woman; I am one\\*
Born of great kings, who walk my ways alone,\\*
\vin Priestess of \textsc{Bast}, the Queen of the Two Worlds,\\*
And seeking no man's gift and fearing none.

If me thou seekest, I will speak within\\*
Mine own house: knock and thou shalt enter in:\\*
\vin In \textsc{Per-Bast}, by the Houses of the Dead,\\*
Past \textsc{Kemi}, where the desert doth begin.

There ask for \textsc{Ta-Buvu\"{e}}. I go hence\\*
Now with my maids to make magnificence\\*
\vin Before thee. And no man shall watch thy way\\*
Into my house, nor mark thy coming thence.'{''}

The Slave returned, and \textsc{Setne}'s heart did fail\\*
For very joy at hearing of the tale.\\*
\vin He called his servants: ``Make me a swift boat\\*
Ready, with rowers and a silken sail.''

And \textsc{Setne} marvelled in his heart a space,\\*
And in his mirror looked; and, lo, his face\\*
\vin Seemed beautiful again, and all his limbs\\*
Light, like a young man when he runs a race.

So walked he to the boat and entered in,\\*
And bade them row as ne'er they rowed, to win\\*
\vin \textsc{Per-Bast}, beyond the Houses of the Dead,\\*
Past \textsc{Kemi}, where the desert doth begin.

And there, behold, a tower exceeding tall\\*
Set in a pleasant place; and a great wall\\*
\vin Was round it, and a garden to the north\\*
With many trees. And \textsc{Setne} gave a call:

``Whose is this tower?'' And heard an answer: ``Here\\*
Dwells \textsc{Ta-Buvu\"{e}}, daughter of the Seer\\*
\vin Of \textsc{Bast}, she who is named The Beautiful.''\\*
And \textsc{Setne} entered and no man was near.

And up the garden ways he went, and cast\\*
His eyes on all and marvelled as he passed:\\*
\vin And \textsc{Ta-Buvu\"{e}} came and held his hand\\*
And spoke: ``Now by the Holiness of \textsc{Bast},

This day is happiness. Come to mine high\\*
Chamber, we two alone amid the sky.''\\*
\vin So up the stair they went, to a cool room\\*
Of turquoise wrought and lapis lazuli.

Couches were there, decked with fair linen strand\\*
Like \textsc{Pharaoh}'s couch; and cups of gold did stand\\*
\vin On a great dresser, and a cup of gold\\*
Was filled with wine and laid in \textsc{Setne}'s hand.

And \textsc{Ta-Buvu\"{e}} said: ``Take wine and meat.''\\*
But he said: ``Love, how could I drink or eat?''\\*
\vin Then in a censer burning gums they brought\\*
And spices rare and unguent for his feet.

Lo, none was like her, none that bore the name\\*
Of woman! And his heart rose like a flame:\\*
\vin ``O \textsc{Ta-Buvu\"{e}}, let the end be now:\\*
Let us make perfect that for which we came!''

``Is not this house thine own and I thy bride?''\\*
Said she: ``Yet one thing first shalt thou decide.\\*
\vin I am no common woman, but to \textsc{Bast}\\*
Pure, and full-filled of majesty and pride.

Thy present wife put from thee, that no soul\\*
Near me may stand; and write it in a scroll.\\*
\vin And all thy wealth of Priesthoods and of Powers,\\*
With me alone thou shalt divide the whole.''

Said \textsc{Setne}: ``Call a scribe.'' And at their call\\*
The scribe came; and he wrote a scroll withal\\*
\vin And cast off \textsc{Setne}'s wife; and \textsc{Setne}'s wealth,\\*
With \textsc{Ta-Buvu\"{e}} did he share it all.

And in that hour one spoke to \textsc{Setne}: ``Lo,\\*
Thy children, waiting in the court below,\\*
\vin Sit with the dogs and curling cats of \textsc{Bast}.''\\*
And \textsc{Setne} said: ``Bring them, and let them know.''

Then \textsc{Ta-Buvu\"{e}} laid her raiment proud\\*
Off, and put on soft linen, like one vowed\\*
\vin To bridal; and her body through the robe\\*
Shone, as the moon shines through a little cloud.

And back she turned to him and poured him wine,\\*
And said: ``These children must not strive with mine.\\*
\vin Make them to sign the scroll, too, and give up\\*
Their part in thee.'' And \textsc{Setne} made them sign.

The fine, fine, linen robed her like a mist\\*
Which robeth \textsc{Ra} in pearl and amethyst;\\*
\vin And \textsc{Setne} marvelled gazing; and again\\*
She spoke, and \textsc{Setne}'s hand she took and kissed:

``These children, knowing all to me thou art,\\*
Hate me.---Let them be mine to take apart\\*
\vin And do my will upon them.'' And he said:\\*
``Do all the abomination of thy heart.''

She slew them then, and from her window fine\\*
Cast them. And far below he heard the whine\\*
\vin Of dogs that tore and curling cats of \textsc{Bast}\\*
Which lapped their blood. And \textsc{Setne} drank his wine.

He said: ``Those children that were slain had birth\\*
By me. O Woman, thou hast made much dearth\\*
\vin About me. Give me that for which I came,\\*
Else have I nothing, nothing, on this earth.''

``Hast thou not me,'' she said, ``in place of all?\\*
Come, therefore!'' And she led him through the hall\\*
\vin To a fair couch, ebon and ivory;\\*
And down he lay, and spread swift arms withal

To clasp her; and within his arms outspread,\\*
Behold, she withered, withered; and her head\\*
\vin It had no eyes, and downward all her jaw\\*
Dropped, like the jaws of the uncared-for dead.

And \textsc{Setne} strove to rise, but cloud on cloud\\*
Held him: hot wind and hate and laughter loud,\\*
\vin And one that wept for a world's glory gone,\\*
And dust, dust, dust: and \textsc{Setne} shrieked aloud:

And saw: and, lo, all naked in the day\\*
In a waste place of bricks and shards he lay,\\*
\vin And clutched a burning kiln. And near him passed\\*
The way and much folk jeering on the way,

Soldiers and priests, beggars and men of pride.\\*
And \textsc{Setne} rolled him in the dust and cried:\\*
\vin ``My children!'' And a great lord rose in wrath:\\*
``Thy children stand this hour at \textsc{Pharaoh}'s side,

Thou naked man! Thou Priest whom none shall bless!\\*
And ask for thee. What? Is it drunkenness?''\\*
\vin And \textsc{Setne} said: ``They live.'' And said: ``O King,\\*
Throw me, I pray, some robe in this distress,

Wrought by dead \textsc{Nefrekepta} in his ire.\\*
I go to yield him up his worst desire,\\*
\vin A fork upon my neck, between my hands\\*
A rod, and on my head a bowl of fire.''

One of the bondmen threw him, at that call,\\*
A poor man's robe; and on to \textsc{Pharaoh}'s Hall\\*
\vin He journeyed with them, and stretched out his arms\\*
And clasped his sons, and told to \textsc{Pharaoh} all.

``Yea, take the Book, take quickly,'' \textsc{Pharaoh} said,\\*
``The rod, the fork, the fire upon thine head,\\*
\vin And seek dead \textsc{Nefrekepta} in his tomb,\\*
And kneel and pray the pardon of the dead.''

And \textsc{Setne} heard; and quick ere set of sun\\*
He stood before the Tomb, and one by one\\*
\vin Passed the great doors, and opened the last door,\\*
And, lo, a light through all the chamber shone,

A great light, like the going forth of \textsc{Ra}.\\*
And while he stood the Woman cried: ``Aha,\\*
\vin \textsc{Setne}, thou com'st! And if thou com'st alive\\*
'Tis \textsc{Ptah} hath saved thee and the grace of \textsc{Ptah}.''

But \textsc{Nefrekepta} laughed. And \textsc{Setne} came\\*
Kneeling: ``O King, with rod and fork and flame\\*
\vin I come,'' he said; ``and yield thee up thy Book.\\*
What is thy judgement? Is it further shame?''

But \textsc{Nefrekepta} laughed: ``I would not now\\*
Make thee my slave, nor smite, nor burn thy brow.\\*
\vin This was enough.---Yet one thing lacketh me\\*
Still, and thereto I bind thee by a vow.

Far off in a strange grave 'mid much annoy\\*
My wife \textsc{Ahure} lieth and the boy\\*
\vin \textsc{Merab}; 'tis but their shadows, by the art\\*
Of a good scribe, dwell here and have no joy.

Therefore I charge upon thee my behest:\\*
Go, bring from \textsc{Coptos} to this House of Rest\\*
\vin My wife \textsc{Ahure} and \textsc{Merab} the boy.''\\*
And \textsc{Setne} rose and took on him the quest.

And straightway before \textsc{Pharaoh} bowed his head\\*
And told him all the tale. And \textsc{Pharaoh} said:\\*
\vin ``I give thee mine own pleasure-ship to sail\\*
To \textsc{Coptos} and bring back those ancient dead.''

So \textsc{Pharaoh}'s pleasure-ship with all its crew\\*
Was brought, and southward on the wind they flew\\*
\vin To \textsc{Coptos}; and the High Priest saw the ship,\\*
And all the Priests, and came in haste thereto.

The Priests of \textsc{Isis} and \textsc{Harpocrates}\\*
And the Chief Priest; \textsc{Setne} to all of these\\*
\vin Gave ox and goose and wine, and with them walked\\*
On \textsc{Coptos} Hill amid the tombs and trees.

Three days and nights among the tombs they trod\\*
In \textsc{Coptos} on the Hill, and every sod\\*
\vin They turned and marked, and every graven stone,\\*
And the Scribes' writings in the House of God.

But never could they find by night nor day\\*
The tomb where \textsc{Merab} and \textsc{Ahure} lay.\\*
\vin And \textsc{Nefrekepta} knew they found it not,\\*
And sent his shadow forth to guide their way.

Like an old man, a bent and aged Priest,\\*
It sate. And \textsc{Setne} said: ``Joy be increased,\\*
\vin O Father! Thou dost know the things of old;\\*
Three days and nights we search, and have not ceased,

To find the tomb which holds \textsc{Ahure} dead\\*
And \textsc{Merab}.'' Then the old man raised his head:\\*
\vin ``The father of my grandsire in old days\\*
Spoke of it to my grandsire; and he said

The father of his grandsire once had told\\*
His grandsire how those two were laid of old\\*
\vin Far in the southmost corner, where the house\\*
Now stands in which the scrivener tells his gold.''

And \textsc{Setne} said: ``Old man, methinks I see\\*
Some hate here. Hath the scrivener injured thee,\\*
\vin That thou wouldst wreck his house and dig beneath?''\\*
He answered: ``Have a watch set over me;

Then raze the scrivener's house, and, under ground\\*
By the south corner, if there be not found\\*
\vin Both \textsc{Merab} and \textsc{Ahure}, have me slain!''\\*
So there they held him and a guard stood round.

The scrivener's house was razed; and that same day\\*
They found where \textsc{Merab} and \textsc{Ahure} lay,\\*
\vin And, like great \textsc{Pharaohs}, down to \textsc{Pharaoh}'s boat\\*
Bore them 'mid Priests and Princes in array.

And \textsc{Setne} sought that ancient man, and, lo,\\*
He was not. By that sign did \textsc{Setne} know\\*
\vin This too was \textsc{Nefrekepta}. Then they built\\*
The scrivener's house again, and turned to go:

And went on board, and back to \textsc{Memphis} bore\\*
Those \textsc{Pharaohs} home, with stream and wind and oar;\\*
\vin Singing they went, and \textsc{Pharaoh} heard them sing;\\*
And \textsc{Pharaoh} rose and met them by the shore,

And led those Mighty Ones in robes of pride\\*
To \textsc{Nefrekepta}'s tomb, and sanctified\\*
\vin Their entering in, and made a mound above;\\*
And there for ever sleep they, side by side.

And there is finished all that fell between\\*
\textsc{Setne} and \textsc{Nefrekepta} and his Queen\\*
\vin \textsc{Ahure} and the boy \textsc{Merab}. 'Twas writ\\*
In the first month of winter, Year \textsc{xv}.
\end{verse}

\end{document}
